---
title: "Class 10: Halloween Mini Project"
author: "Ian Gurholt (PID:A16767484)"
format: gfm
---

##Importing Data

```{r}
candy_file <- "candy-data.csv"


candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
There are 85 different types of candy in the data set, each with their own row and found using nrow() function.

Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 types of fruity candy in the data set by suming up all of the fruity column using the `sum()` function.

##Favorite Candy

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Snickers", ]$winpercent
```
My favorite candy is Snickers and its win percentage is 76,67%

>Q. Can you find all the candies with a win percent above 50 that are fruity?

```{r}
library(dplyr)

candy |>
  filter(winpercent>50) |>
  filter(fruity==1)
```

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
Win percentage for kit kat candy is 76.77%

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bar", ]$winpercent
```
Win percentage for Tootsie Roll is 49.65%

```{r}
library("skimr")
skim(candy)
```
>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Looks like the win percent variable/column is measured on a different scale than everything else! I will need to scale my data before conducting any analysis such as PCA etc.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A one represents a candy that is classified as chocolate while a zero represents a candy that is not classified as chocolate

> Q8. Plot a histogram of winpercent values

We can do this a few ways, the "base" R `hist()` function or with gg plot
```{r}
hist(candy$winpercent, breaks = 30)
```
```{r}
library(ggplot2)
ggplot(candy)+ 
  aes(x= winpercent)+
  geom_histogram(binwidth=8)
```

> Q9. Is the distribution of winpercent values symmetrical?

No, according to the histograms, there appears to be a skew to the left, with a right hand tail based on the data

>Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```
It is below 50% as we utilize the median to access the center of the distribution, which according to the summary is less than 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
fruit.candy<- candy |>
  filter(fruity==1)

summary(fruit.candy$winpercent)
```
```{r}
choc.candy<- candy |>
  filter(chocolate==1)

summary(choc.candy$winpercent)
```
On average, chocolate candy is higher ranked as the median winpercent score of 60.8 is greater than fruity candy's 42.97

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.candy$winpercent, fruit.candy$winpercent)
```
Yes this difference is statistically significant as p-value is much less than 0.05 threshold value

> Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),],5)
```
These are the 5 least liked candies in this data set.

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
play<- c("d","a","c")
sort(play)
order(play)
```
```{r}
head(candy[order(candy$winpercent, decreasing = T),],5)
```
These are the top 5 most liked candies

>Q15. Make a first barplot of candy ranking based on winpercent values.

Let's do a barplot of winpercent values
```{r}
ggplot(candy)+
  aes(winpercent, row.names(candy))+
  geom_col()
```
> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy)+
  aes(winpercent, reorder(row.names(candy), winpercent), fill=chocolate)+
  geom_col()
```
I want a more customized color scheme where I can see both chocolate and bar and fruit from a single plot, without having to make multiple plots. To do this we can roll our own color vector.

```{r}
mycols<- rep("black", nrow(candy))
mycols[as.logical(candy$chocolate)]<- "chocolate"
mycols[as.logical(candy$bar)]<- "brown"
mycols[as.logical(candy$fruity)]<- "pink"
mycols[rownames(candy)=="Snickers"]<- "blue"
```
```{r}
ggplot(candy)+
  aes(winpercent, reorder(row.names(candy), winpercent))+
  geom_col(fill=mycols)
```
> Q17. What is the worst ranked chocolate candy?

According to the plot, the worst ranked chocolate candy is sixlets

> Q18. What is the best ranked fruity candy?

According to the plot, the best ranked fruity candy is starburst

Plot of Win Percent versus Price Percent to see what would be the best candy to buy...

```{r}
mycols[as.logical(candy$fruity)]<- "red"
```

```{r}
ggplot(candy)+
  aes(winpercent, pricepercent)+
  geom_point(col=mycols)
```
Add labels
```{r}
ggplot(candy)+
  aes(winpercent, pricepercent, label=rownames(candy))+
  geom_point(col=mycols)+
  geom_label()
  
```
```{r}
ggplot(candy)+
  aes(winpercent, pricepercent, label=rownames(candy))+
  geom_point(col=mycols)+
  geom_text(col=mycols)
```
Make Labels non-overlapping
```{r}
library(ggrepel)
ggplot(candy)+
  aes(winpercent, pricepercent, label=rownames(candy))+
  geom_point(col=mycols)+
  geom_text_repel(col=mycols, max.overlaps=8)
```
> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reeses minatures are ranked highest in terms of winpercent for the least money according to the plot above

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The top 5 most expensive candy types are Nik L Nip, Ring pops, Smarties, Hersheys Krackel, and Hersheys chocolate, with Nik L Nip being the least popular of them all.

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```


##Exploring the Correlation Structure

```{r}
library(corrplot)
```

```{r}
cij<- cor(candy)
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are anticorrelated based on the correlation structure

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and win percentage are most positively correlated based on the correlation structure

##Principal Component Analysis

```{r}
pca<- prcomp(candy, scale=T)
summary(pca)
```
```{r}
plot(pca$x[,1],pca$x[,2], col=mycols, pch=16)
```
How do the original columns contribute to the new PCs. I will look at PC1 first here

```{r}
loadings<- as.data.frame(pca$rotation)

ggplot(loadings)+
  aes(PC1, reorder(rownames(loadings), PC1), fill=PC1) +
  geom_col()
```
##ggplot of PCA 

```{r}
my_data <- cbind(candy, pca$x[,1:3]) 
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=mycols)

p
```
```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

PC1 strongly picks up fruity, hard, and pluribus in the positive direction as these variables are all positioned on the positive right side of PC1 loadings plot compared to all other variables being positioned on the left, negative side of the loadings plot. This makes sense to me as naturally, hard, fruity, candy are commonly found related to each other as well as being pluribus (bagged), and their close relationship via the PC1 plot support this real world finding statistically. 




